---
title: "COVID-19 in Michigan"
author: "Clifton Franklund"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---
```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyquant)
library(ggpubr)
library(stringr)
library(readxl)
library(choroplethr)
library(choroplethrMaps)
library(RColorBrewer)
library(tadaatoolbox)
library(scales)
```

```{r data, echo=FALSE, message=FALSE, warning=FALSE}

# Load COVID-19 case data from Michigan state site
download.file("https://www.michigan.gov/documents/coronavirus/Cases_by_County_and_Date_2020-09-26_703515_7.xlsx", destfile="./rawData/Cases_by_County_and_Date.xlsx")

michiganData <- read_xlsx("./rawData/Cases_by_County_and_Date.xlsx")

# Load county census data
censusData <- read_csv("./rawData/csvData.csv")

# Load county map codes
data(county.regions)

# Load population densities
density <- read_csv("./rawData/density.csv")
```

```{r cleanData, echo=FALSE, message=FALSE, warning=FALSE}

# Clean up census data set
censusData <- censusData %>% 
  rename(County = CTYNAME) %>% 
  mutate(County = str_remove_all(County," County"))
censusData$County <- str_to_lower(censusData$County)

# Lower case county names
density$County <- str_to_lower(density$County)

# Get just Michigan county names in map data
county.regions <- filter(county.regions, state.name == "michigan")

## Clean up michiganData column names
michiganData <- michiganData %>% 
  rename(County = COUNTY, Status = CASE_STATUS) %>% 
  filter(County != "MDOC", County != "Out-of-State", County != "FCI", County != "Unknown", !is.na(Date))
michiganData$County <- str_to_lower(michiganData$County)
michiganData$County <- recode(michiganData$County, "st clair" = "st. clair", "st joseph" = "st. joseph")

# Combine Detroit City with Wayne County
detroit <- michiganData %>% filter(County == "detroit city")
wayne <- michiganData %>% filter(County == "wayne")
combined <- wayne
combined$Cases <- combined$Cases + detroit$Cases
combined$Cases.Cumulative <- combined$Cases.Cumulative + detroit$Cases.Cumulative
combined$Deaths <- combined$Deaths + detroit$Deaths
combined$Deaths.Cumulative <- combined$Deaths.Cumulative + detroit$Deaths.Cumulative
michiganData <- subset(michiganData, County != "detroit city")
michiganData <- subset(michiganData, County != "wayne")
michiganData <- rbind(michiganData, combined)

# Add mapping data to the dataset
michiganData <- left_join(michiganData, county.regions, by =c("County" = "county.name"))

# Add census data to the datset
michiganData <- left_join(michiganData, censusData, by = c("County"))

# Add population densities
michiganData <- left_join(michiganData, density, by = c("County"))

# Get just the confirmed cases
michiganConfirmedData <- michiganData %>% 
  filter(Status == "Confirmed")
write_csv(michiganConfirmedData, "./processedData/michiganConfirmedData.csv")

# Get just the probable cases
michiganProbableData <- michiganData %>% 
  filter(Status == "Probable")
write_csv(michiganProbableData, "./processedData/michiganProbableData.csv")

# Add confirmed and probable cases to get total cases
michiganTotalData <- tibble(County = michiganConfirmedData$County, Date = michiganConfirmedData$Date, Status = "Total", Cases = michiganConfirmedData$Cases + michiganProbableData$Cases, Deaths = michiganConfirmedData$Deaths + michiganProbableData$Deaths, Cases.Cumulative = michiganConfirmedData$Cases.Cumulative + michiganProbableData$Cases.Cumulative, Deaths.Cumulative = michiganConfirmedData$Deaths.Cumulative + michiganProbableData$Deaths.Cumulative, Updated = michiganConfirmedData$Updated, Region = michiganConfirmedData$region, County.FIPS.Character = michiganConfirmedData$county.fips.character, State.Name = michiganConfirmedData$state.name, State.FIPS.Character = michiganConfirmedData$state.fips.character, State.Abb = michiganConfirmedData$state.abb, pop2018 = michiganConfirmedData$pop2018, GrowthRate = michiganConfirmedData$GrowthRate, Density = michiganConfirmedData$Density)
metroCounty <- c("macomb", "oakland", "wayne")
michiganTotalData$Detroit <- ifelse(michiganTotalData$County %in% metroCounty, "Yes", "No")

write_csv(michiganTotalData, "./processedData/michiganTotalData.csv")

# Summarize total numbers for the three-county metro Detroit area
metroDetroit <- michiganTotalData %>% 
  filter(County == "wayne" | County == "oakland" | County == "macomb") %>% 
  group_by(Date) %>% 
  mutate(Cases = sum(Cases), Deaths = sum(Deaths), Cases.Cumulative = sum(Cases.Cumulative), Deaths.Cumulative = sum(Deaths.Cumulative))
metroDetroit <- metroDetroit %>% filter(County == "wayne")
metroDetroit$County <- "Metro Detroit"
write_csv(metroDetroit, "./processedData/metroDetroit.csv")

# Summarize total numbers for the rest of Michigan
notDetroit <- michiganTotalData %>% 
  filter(County != "wayne" & County != "oakland" & County != "macomb") %>% 
  group_by(Date) %>% 
  mutate(Cases = sum(Cases), Deaths = sum(Deaths), Cases.Cumulative = sum(Cases.Cumulative), Deaths.Cumulative = sum(Deaths.Cumulative))
notDetroit <- notDetroit %>% filter(County == "mecosta")
notDetroit$County <- "Not Detroit"
write_csv(notDetroit, "./processedData/notDetroit.csv")

# Summarize total numbers for the entire state of Michigan
total <- michiganTotalData %>% 
  group_by(Date) %>% 
  mutate(Cases = sum(Cases), Deaths = sum(Deaths), Cases.Cumulative = sum(Cases.Cumulative), Deaths.Cumulative = sum(Deaths.Cumulative))
total$County <- "Total"
write_csv(total, "./processedData/total.csv")

# Create on comparison set
compareDetroit <- rbind(total, metroDetroit, notDetroit)
compareDetroit$Date <- as.POSIXct(as.Date(compareDetroit$Date))

metroDataCount <- michiganTotalData %>% 
  group_by(County) %>% 
  summarize(Cases = max(Cases.Cumulative), Deaths = max(Deaths.Cumulative), Detroit = Detroit)
metroDataCount <- unique(metroDataCount)

metroDataNorm <- michiganTotalData %>% 
  group_by(County) %>% 
  summarize(Cases = max(Cases.Cumulative)/pop2018*100000, Deaths = max(Deaths.Cumulative)/pop2018*100000, Detroit = Detroit)
metroDataNorm <- unique(metroDataNorm)

weekly <- michiganTotalData %>% group_by(County, Week = week(Date)) %>% summarise(Weekly.Cases = sum(Cases), Weekly.Deaths = sum(Deaths), Case.Mortality=Weekly.Deaths/Weekly.Cases*100)

dataSummary <- michiganTotalData %>% 
  group_by(Region) %>% 
  rename(region = Region) %>% 
  summarize(value = max(Deaths.Cumulative), pop = pop2018)
dataSummary <- unique(dataSummary)

```

## Abstract  
This is an exploratory analysis of some COVID-19 trends in Michigan that interest me. I strongly suspect that the vast majority of cases and adverse outcomes are concentrated in the Detroit area. Our response as a state, however, has not really been defined geographically. This is probably not the best way to steward our limited resources (note for future microbial pandemics). I am also keenly interested in the impact that reopening college campuses will have on the overall trajectory of COVID-19 incidence in our region. You can [view this report](https://reproducibleassessment.com/Michigan-COVID/index.html) online. The data and all analyses performed can be accessed on [my GitHub repo](https://www.github.com/WeeBeasties/Michigan-COVID). Feel free to fork and use my repo.

## Data  
One should always be a bit skeptical of analyses with pretty graphs and figures. It is important to know what data were used, how they were manipulated, and also the trustworthyness of the data sources. TheMichigan COVID-19 data for this document were obtained at https://www.michigan.gov/coronavirus/0,9753,7-406-98163_98173---,00.html in the Public Use Datasets called "Cases by County and Date". The **September 26, 2020** dataset was used to create this report. Michigan county populations were found at https://worldpopulationreview.com/us-counties/states/mi which gives the 2018 census numbers. These were the most recent reliable numbers that I could find at the moment.

## Data munging
The Michigan COVID-19 dataset had a few peculiarities that I need to deal with in order to analyze it easily. There were several "counties" that are not actual Michigan counties that needed to be cleaned. The values for "Out-of-State" are not relevant for my purposes and were removed from the analyses. Likewise, prison populations ("FCI" and "MDOC") cannot be ascribed to particular counties and were, therefore, also removed. Cases with "Unknown" origins were also removed. Finally, the dataset lists "City of Detroit" as a separate entity, those numbers were added back into it's home country (Wayne) to facilitate analysis.

The number of cases and deaths per day are recorded for both confirmed and probable cases. To be as conservative as possible, these numbers were added together to create a total number of cases and deaths per day from each county (along with total cumulative cases and deaths). These total values, confirmed + probable, are used for all of the analyses in this report.

GIS information was left-joined onto the total data. The census data was also left-joined and used to create population-normalized values (cases or deaths per 100,000 population).

## Statewide distribution   
Many websites report COVID-19 cases or deaths in Michigan using county maps. The Michigan state map can be found [here](https://www.michigan.gov/coronavirus/0,9753,7-406-98163_98173---,00.html). This representation can be a little misleading, though. One might assume that the intensity of the colors is proportional to the number of cases or deaths - this is not necessarily true. If you examine the legends you will notice that the color bins are not necessarily spaced out evenly. Figure 1 shows the total number of cases and deaths by county in Michigan using a linear color scale.

```{r state_map1, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, fig.cap="**Figure 1:** State-wide distribution of COVID-19. Panel A shows total cases, while panel B shows total deaths."}
mapDataA <- michiganTotalData %>% 
  group_by(Region) %>% 
  rename(region = Region) %>% 
  summarize(value = max(Cases.Cumulative))
mapDataA <- unique(mapDataA)

mapA <- county_choropleth(mapDataA, state_zoom = "michigan", num_colors = 1)

mapDataB <- michiganTotalData %>% 
  group_by(Region) %>% 
  rename(region = Region) %>% 
  summarize(value = max(Deaths.Cumulative))
mapDataB <- unique(mapDataB)

mapB <- county_choropleth(mapDataB, state_zoom = "michigan", num_colors = 1) 

ggarrange(mapA, mapB, labels = c("A", "B"), ncol = 1, nrow = 2, font.label = list(size = 28, color = "black", face = "bold", family = NULL))
```

As expected, the metropolitan Detroit tri-county (Wayne, Oakland, and Macomb) area has the vast majority of reported cases and deaths. The exact values are summarize in Table 1.

```{r countTable, echo=FALSE, message=FALSE, warning=FALSE}
dataTable <- metroDataCount %>% 
  group_by(Detroit) %>% 
  summarize(Total.Cases = sum(Cases), Total.Deaths = sum(Deaths))
knitr::kable(dataTable, format.args = list(big.mark = ",", digits = 1), caption = "**Table1:** Total number of COVID-19 cases and deaths in metopolitan Detroit (Wayne, Oakland, and Macomb counties) vs all other Michigan counties combined.")
```

The three-county metropolitan Detroit area accounted for `r format(dataTable$Total.Cases[2], big.mark=",", scientific=FALSE)` out of `r format(dataTable[1,2] + dataTable[2,2], big.mark=",", scientific=FALSE)` total cases in Michigan (`r round(dataTable[2,2]/(dataTable[1,2] + dataTable[2,2])*100,1)`%). Metropolitan Detroit also represented `r format(dataTable$Total.Deaths[2], big.mark=",", scientific=FALSE)` out of `r format(dataTable[1,3] + dataTable[2,3], big.mark=",", scientific=FALSE)` total COVID-19 deaths in Michigan (`r round(dataTable[2,3]/(dataTable[1,3] + dataTable[2,3])*100,1)`%).

But wait, you say, don't most people live in those counties? Well... not _most_, but quite a few; let's normalize the cases and deaths by county population and try that again.

```{r state_map2, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=12, fig.cap="**Figure 2:** State-wide distribution of COVID-19 normalized for county population. Panel A shows total cases per 100,000 people, while panel B shows total deaths per 100,000 people."}
mapDataC <- michiganTotalData %>% 
  group_by(Region) %>% 
  rename(region = Region) %>% 
  summarize(value = max(Cases.Cumulative)/pop2018*100000)
mapDataC <- unique(mapDataC)
mapC <- county_choropleth(mapDataC, state_zoom = "michigan", num_colors = 1)

mapDataD <- michiganTotalData %>% 
  group_by(Region) %>% 
  rename(region = Region) %>% 
  summarize(value = max(Deaths.Cumulative)/pop2018*100000)
mapDataD <- unique(mapDataD)
mapD <- county_choropleth(mapDataD, state_zoom = "michigan", num_colors = 1)

ggarrange(mapC, mapD, labels = c("A", "B"), ncol = 1, nrow = 2, font.label = list(size = 28, color = "black", face = "bold", family = NULL))
```

Ok, so Figure 2 shows that when adjusting for population the number of cases is more widespread across the state. The number of cases still appears to be greater below a latitude of about 43.5°N. The number of deaths per 100,000 is interesting because you can discern the interstate highway system on the map. the I-75 corridor is a fairly dark blue color. The I-94 and I-96 corridors are also visible, but less blue. This is not necessarily causative (most urban centers are arrayed along the interstate highways). Table 2 shows the differences between Detroit and the rest of the state.

```{r norm_table, echo=FALSE, message=FALSE, warning=FALSE}
normDataTable <- metroDataNorm %>% 
  group_by(Detroit) %>% 
  summarize(Mean.Cases = mean(Cases), SD.Cases = sd(Cases), Mean.Deaths = mean(Deaths), SD.Deaths = sd(Deaths))
knitr::kable(normDataTable, format.args = list(big.mark = ",", digits = 2), caption = "**Table2:** Total number of COVID-19 cases and deaths per 100,000 people in metopolitan Detroit (Wayne, Oakland, and Macomb counties) vs all other Michigan counties combined.")
my_t.test <- t.test(Deaths~Detroit, data=metroDataNorm)
```
After normalizing for county population, metropolitan Detroit counties average `r round(normDataTable[2,2]/normDataTable[1,2],1)`-times more COVID-19 cases and `r round(normDataTable[2,4]/normDataTable[1,4],1)`-times more COVID-19 deaths than the rest of Michigan counties. On average, metropolitan Detroit (m = `r round(my_t.test$estimate[2],1)`) experienced significantly higher rates of COVID-19 deaths per 100,000 people than the rest of the state (m = `r round(my_t.test$estimate[1],1)`), *t*(`r round(my_t.test$parameter[1],2)`) = `r round(my_t.test$statistic[1],2)`, *p* < `r round(my_t.test$p.value,3)`.

```{r by_county, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=12, fig.cap="**Figure 3:** Normalized COVID-19 death rates by county (deaths per 100,000 people). The vertical line at 9.7 people per 100,000 is the annual traffic fatality rate for Michigan."}

ggplot(metroDataNorm, aes(x=reorder(County, +Deaths), y=Deaths)) + 
  geom_point(col="firebrick", size=3) +   # Draw points
  geom_segment(aes(x=County, 
                   xend=County, 
                   y=min(Deaths), 
                   yend=max(Deaths)), 
               linetype="dashed", 
               size=0.1) +
  ylab("Deaths per 100,000 population") +
  xlab("County") +
  geom_hline(yintercept = 9.7, linetype=1) +
  coord_flip() +
  theme_bw()
```

Figure 3 highlights the fact that deaths per capita is not equal across all counties. Seven counties (including the Detroit metropolitan area) and much higher. Nearly three dozen counties have normalized COVID-19 death rates lower than that of car fatalities in the state. A total of `r prettyNum(sum(dataSummary$value),big.mark=",",scientific=FALSE)` people have died of COVID-19 out of a population of `r prettyNum(sum(dataSummary$pop),big.mark=",",scientific=FALSE)` people. That gives a normalized value of `r round(sum(dataSummary$value)/sum(dataSummary$pop)*100000,1)` people per 100,000.

## Detroit vs everyone else

Since the Detroit metropolitan area seems to be hit harder than the rest of the state, I wondered what the trends of cases and deaths looked like over time. Figure 4 shows the trends of total cases and deaths along with a five-day moving average of each. 

```{r Detroit_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=10, fig.cap="**Figure 4:** A comparison of COVID-19 incidence between metropolitan Detroit (red) and the rest of Michigan (blue). The total state numbers are shown in green. Seven-day moving averages of the data are shown as dotted lines. The veritcal lines denote holidays: The 4th of July and Labor Day on September 7th."}

plotA <- ggplot(compareDetroit) +
  geom_line(aes(x=Date, y=Cases, color=County)) +
  geom_rect(data = data.frame(xmin=as.POSIXct(as.Date(c("2020-07-01"))),xmax=as.POSIXct(as.Date(c("2020-07-15"))),ymin = -Inf,ymax = Inf),aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),fill = "grey", alpha = 0.3) +
  ylim(0,1700) +
  scale_x_datetime(breaks = date_breaks("months"),labels = date_format("%b")) +
  annotate(geom="text",x=as.POSIXct(as.Date("2020-07-07")),
    y=1650,label="July 4th\nHoliday",fontface="bold") +
    geom_rect(data=data.frame(xmin=as.POSIXct(as.Date(c("2020-08-28"))),xmax=as.POSIXct(as.Date(c("2020-09-24"))),ymin = -Inf,ymax = Inf),aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),fill = "grey", alpha = 0.3) +
  annotate(geom="text",x=as.POSIXct(as.Date("2020-09-11")),
    y=1650,label="Universities\nReopen",fontface="bold") +
  ylim(0,1700) +
  scale_colour_manual(values=c('Total'="#336600", 'Metro Detroit'="firebrick", 'Not Detroit'="dodgerblue")) +
  geom_ma(aes(x=Date, y=Cases, color=County),ma_fun = EMA, n = 7, wilder = TRUE, linetype = 3, size = 1.25) +
  geom_vline(xintercept = as.POSIXct(as.Date(c("2020-07-04", "2020-09-07"))), linetype=4) +
  theme_bw() +
  theme(legend.position = "none") +
  theme(axis.text=element_text(size=14),
        axis.title=element_text(size=18,face="bold"))

plotB <- ggplot(compareDetroit, aes(x=Date, y=Deaths, color=County)) +
  geom_line() +
  scale_x_datetime(breaks = date_breaks("months"),labels = date_format("%b")) +
  ylim(0,170) +
  scale_colour_manual(values=c('Total'="#336600", 'Metro Detroit'="firebrick", 'Not Detroit'="dodgerblue")) +
  geom_ma(ma_fun = EMA, n = 7, wilder = TRUE, linetype = 3, size = 1.25) +
  geom_vline(xintercept = as.POSIXct(as.Date(c("2020-07-04", "2020-09-07"))), linetype=4) +
  theme_bw() +
  theme(legend.position = "none") +
    theme(axis.text=element_text(size=14),
        axis.title=element_text(size=18,face="bold"))

ggarrange(plotA, plotB, labels = c("A", "B"), ncol = 1, nrow = 2, font.label = list(size = 20, color = "black", face = "bold", family = NULL))
```

The intial spike of COVID-19 cases appears to have spiked around the beginning of April in Detroit and about a month later in other counties (abeit at a much lower level). As restrictions were dropped in Michigan in June, the number of cases began to increase again. They appear to be flattening out over the past month. Interestingly, The majority of all COVID-19 deaths in Michigan occurred in Detroit in the first wave of infections. The reemergence of COVID-19 cases through July and August has not been accompanied by a coincident increase in deaths. Perhaps less severe cases are being detected more frequently now due to increased testing. There are two periods of time (shaded gray) where the number of cases from Detroit are lower than those for the rest of the state. Those represent the fourth of July holiday (probably poor social distancing) and, more recently, the reopening of universities in the state. If the school numbers can be brought into check, the number of cases statewide could theortetically be cut in half.

## Comparing Isabella (Central Michigan University), Mecosta (Ferris State University) and Ottawa (Grand Valley State University) county cases  

Finally, I am also interested in comparing how Ferris State University does in controlling COVID-19 compared to Central Michigan University and Grand Valley State University. Figure 5 compares the total number of COVID-19 cases between those three counties.

```{r county_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7, fig.cap="**Figure 5:** A comparison of the number of total cases in Isabella, Mecosta, and Ottawa counties. The bars indicate total daily cases and the lines are seven-day moving averages. The vertical lines indicate the fourth of July, CMU opening on August 17th, and GVSU and FSU opening on August 31st, and Labor Day on September 7th."}
counties <- michiganTotalData %>% 
  filter(County == "isabella" | County == "mecosta" | County == "ottawa")
ggplot(counties, aes(x=Date, y=Cases, color=County, fill=County)) +
  scale_color_manual(values=c('mecosta'="firebrick", 'isabella'="dodgerblue", 'ottawa'="steelblue")) +
  scale_fill_manual(legend, values=c('mecosta'="firebrick", 'isabella'="dodgerblue", 'ottawa'="steelblue")) +
  scale_x_datetime(breaks = date_breaks("months"),labels = date_format("%b")) +
  geom_bar(stat = "identity") +
  geom_ma(ma_fun = EMA, n = 7, wilder = TRUE, linetype = 1, size = 2) +
  geom_vline(xintercept = as.POSIXct(as.Date(c("2020-07-04", "2020-08-17", "2020-08-31", "2020-09-07"))), linetype=4) +
  facet_grid(rows=vars(County)) +
  theme_bw() +
  theme(legend.position="none") +
    theme(axis.text=element_text(size=14),
        axis.title=element_text(size=18,face="bold"))
```

```{r weekly_county_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7, fig.cap="**Figure 6:** A comparison of the number of total cases in Isabella, Mecosta, and Ottawa counties by week. The bars indicate total daily cases and the lines are four-week moving averages." }
counties <- weekly %>% 
  filter(County == "isabella" | County == "mecosta" | County == "ottawa")
ggplot(counties, aes(x=Week, y=Weekly.Cases, color=County, fill=County)) +
    scale_color_manual(values=c('mecosta'="firebrick", 'isabella'="dodgerblue", 'ottawa'="steelblue")) +
  scale_fill_manual(legend, values=c('mecosta'="firebrick", 'isabella'="dodgerblue", 'ottawa'="steelblue")) +
  geom_bar(stat = "identity") +
  geom_ma(ma_fun = EMA, n = 4, wilder = TRUE, linetype = 1, size = 2) +
  facet_grid(rows=vars(County)) +
  theme_bw() +
  theme(legend.position="none") +
    theme(axis.text=element_text(size=14),
        axis.title=element_text(size=18,face="bold"))
```

CMU students went back to school earlier than Ferris students. It appears that this influx of students (and increased testing on campus) resulted in a spike of cases. It will be interesting to see how big of a spike Mecosta county experiences over the next few weeks. CMU is twice as large as Ferris, but has 10-times more COVID-19 cases. This probably reflects the fact that they recruit more strongly from counties with a higher endemic COVID-19 rate (near Detroit). Ferris, on the other hand, is primarily a regional institution and the counties that it draws most of its students from have lower overall COVID-19 infection rates. This is one time when living in the sticks has tangible upsides. Here is Ferris by itself to expand the y-axis a little bit.

```{r ferris_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=7, fig.cap="**Figure 7:** Mecosta cases (which includes Ferris State University). The bars indicate total daily cases and the lines are seven-day moving averages. The vertical lines indicate the fourth of July, FSU opening on August 31st, and Labor Day."}
mecosta_county <- michiganTotalData %>% 
  filter(County == "mecosta")
ggplot(mecosta_county, aes(x=Date, y=Cases, color=County, fill=County)) +
  scale_color_manual(values=c('mecosta'="firebrick", 'isabella'="dodgerblue", 'ottawa'="steelblue")) +
  scale_fill_manual(legend, values=c('mecosta'="firebrick", 'isabella'="dodgerblue", 'ottawa'="steelblue")) +
  scale_x_datetime(breaks = date_breaks("months"),labels = date_format("%b")) +
  geom_bar(stat = "identity") +
  geom_ma(ma_fun = EMA, n = 7, wilder = TRUE, linetype = 1, size = 2) +
  geom_vline(xintercept = as.POSIXct(as.Date(c("2020-07-04","2020-08-31", "2020-09-07"))), linetype=4) +
  theme_bw() +
  theme(legend.position="none") +
    theme(axis.text=element_text(size=14),
        axis.title=element_text(size=18,face="bold"))
```

For comparision sake, here are the numbers of positive COVID-19 cases being reported on the [Ferris COVID dashboard page](https://www.ferris.edu/HTMLS/news/coronavirus/covid-19-dashboard.htm).

```{r dashboard, echo=FALSE, comment=FALSE, error=FALSE, warning=FALSE, fig.width=12, fig.height=8, fig.cap="**Figure 8:** Total number of COVID-19 cases. Ferris reports its numbers every Thursday and began collecting data on August 21. The state-reported numbers for Mecosta county are shown for the same intervals."}
dates <- c("08/24-8/27", "08/28-09/03", "09/04-09/10","9/11-9/17","9/18-9/24","9/25-10/2")
week1 <- as.integer(michiganTotalData %>%  filter(County == "mecosta", Date >= "2020-08-24" & Date <= "2020-08-27") %>% summarize(sum(Cases)))
week2 <- as.integer(michiganTotalData %>%  filter(County == "mecosta", Date >= "2020-08-28" & Date <= "2020-09-03") %>% summarize(sum(Cases)))
week3 <- as.integer(michiganTotalData %>%  filter(County == "mecosta", Date >= "2020-09-04" & Date <= "2020-09-10") %>% summarize(sum(Cases)))
week4 <- as.integer(michiganTotalData %>%  filter(County == "mecosta", Date >= "2020-09-11" & Date <= "2020-09-17") %>% summarize(sum(Cases)))
week5 <- as.integer(michiganTotalData %>%  filter(County == "mecosta", Date >= "2020-09-18" & Date <= "2020-09-24") %>% summarize(sum(Cases)))
week6 <- as.integer(michiganTotalData %>%  filter(County == "mecosta", Date >= "2020-09-25" & Date <= "2020-10-02") %>% summarize(sum(Cases)))
ferris <- c(4,19,34,45,18,5)
mecosta <- c(week1,week2,week3,week4,week5,week6)
dataSet <- tibble(Date = dates, Ferris = ferris, Mecosta = mecosta) 

plotData <- dataSet %>%
  pivot_longer(!Date, names_to = "Source", values_to = "Cases")

ggplot(plotData, aes(x=Date, y=Cases, fill=Source)) +
  geom_bar(stat='identity', position='dodge', color='black') +
  scale_fill_manual(values=c("firebrick","gold")) +
  theme_bw() +
  theme(axis.text=element_text(size=14),
        axis.title=element_text(size=18,face="bold"))
```

As of September 17, Ferris State accounts for the vast majority of all reported cases in Mecosta county. There does not yet appear to be a community-associated outbreak at this point. The Mecosta county numbers often take several days to get processed and reported. It is sort of weird, though, that the Ferris numbers no longer seem to be reflected in the county data. We will see if the numbers now begin to decline or if COVID-19 moves from campus out into the general population.

```{r ferris, echo=FALSE, comment=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=7, fig.cap="**Figure 9:** Total COVID-19 cases at Ferris State University since the beginning of the fall semester."}

ferris_days <- c(4,11,18,22,23,24,25,26,29,30,31,32,33)
ferris_cases <- c(4,19,53,82,84,88,98,105,113,114,116,116,121)
ferris_total <- tibble(Days=ferris_days, Cases=ferris_cases)

fit <- glm(Cases ~ Days, data=ferris_total, family=poisson())

ggplot(ferris_total, aes(x=Days, y=Cases)) +
  geom_point(size=3, color="black", fill="firebrick") +
  stat_smooth(method="loess", se=TRUE,  color="firebrick", fill="firebrick", alpha=0.3) +
  scale_y_log10(limits = c(1,500)) +
  annotation_logticks(sides = "l") +
  ylab("Log(Total COVID-19 Cases)") +
  xlab("Time Since Fall Semester Began (Days)") +
  theme_bw() +
  theme(axis.text=element_text(size=14),
        axis.title=element_text(size=18,face="bold"))
```

There appears to have been an exponential increase of COVID-19 cases on campus for the first few weeks of classes. A few positive students caused an initial outbreak within the dorms. The past few days seems to show a little slower spread. Perhaps we are beginning to bend the curve a bit - time will tell.

## Conclusions  

* COVID-19 cases and deaths are not uniformly distributed across the state.
* Metopolitan Detroit accounts for about 56% of all COVID-19 cases and 75% of all deaths.
* After adjusting for county population, metropolitan Detroit counties have nearly 3-times more cases and 7-times more deaths than other Michigan countiues.
* The difference between metropolitan Detroit and other counties is statistically significant.
* The return of students to campus will coincide with an intial spike in COVID-19 cases. It will be interesting to see how well this spike can be controlled on campus. Campus appears to have had an exponential outbreak (albeit small) for the first few weeks. The cases appear to be plateauing after that.


